---
title: "curtis1998_tree"
author: "ML"
date: "25/10/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("devtools","fulltxt")
#library(readxl)
update.packages("fulltxt")
library(fulltext)
#devtools::install_github("ropensci/rotl", dependencies = TRUE, build_vignette=TRUE)
#install.packages("rotl")
library(rotl)
library(metafor)
library(ape)

rm(list=ls())
sessionInfo() #R version 3.2.4 (2016-03-10)
options(scipen=100)
```

For the dataset dat.curtis1998 (included in the metafor package) create tree using rotl - an R package (https://peerj.com/preprints/1471/) to access synthetic phylogenetic tree available at Open Tree of Life database (https://opentree.wikispaces.com/). Note: crashes on some species lists.

```{r load data}
dat <- get(data(dat.curtis1998))
str(dat) # 102 rows
head(dat)
tail(dat)
```

#The function tnrs_match_names returns a data frame that lists the Open Tree identifiers as
#well as other information to help users ensure that the taxa matched are the correct ones

```{r create species list}
species <- paste(dat$genus, dat$species, sep=" ") 
str(species)
species <- unique(tolower(species))
length(species) #37
species #note: "populusx euramericana" should be same as "populus euramericana"
species <- gsub("populusx euramericana", "populus euramericana", species)
species <- unique(tolower(species))
length(species) #36
```

```{r match species list to otl taxa list}
taxa <- tnrs_match_names(names = species)
head(taxa)
dim(taxa) #36
```

#Now that the taxon names are matched to the Open Tree identifiers, we can pass them to the
#function tol_induced_subtree to retrieve the relationships among these taxa. In turn, the tree
#can be plotted directly as it is returned as an ape::phylo object (Figure1).

```{r get tree for taxa}
tree <- tol_induced_subtree(ott_ids= taxa[["ott_id"]]) 
#The following OTT ids were not found: [3915043].
taxa[taxa$ott_id==3915043,] #populus euramericana Populus x canadensis             FALSE 3915043       TRUE HYBRID
#note: Populus × canadensis, known as Canadian poplar or Carolina poplar, is a naturally occurring hybrid of Populus nigra and Populus deltoides
```

```{r fix species list}
populus <- tnrs_match_names(names = ("populus deltoides"))
populus
#replace populus euramericana with populus deltoides (note 2 versions of the name)
species <- gsub("populus euramericana", "populus deltoides", species)
```

```{r get tree again}
taxa <- tnrs_match_names(names = species)
head(taxa)
dim(taxa) #36

tree <- tol_induced_subtree(ott_ids= taxa[["ott_id"]]) 
str(tree)
```

```{r plot tree }
plot(tree, cex=.8, label.offset =.1, no.margin = TRUE)
```

```{r strip ott numbers}
tree$tip.label <- gsub("\\_ott.*","", tree$tip.label)
tree$tip.label <- gsub("_"," ", tree$tip.label)
tree$tip.label
```

```{r replace back populus hybrid }
#replace populus euramericana with populus deltoides
tree$tip.label <- gsub( "Populus deltoides", "Populus euramericana", tree$tip.label)
```

```{r check tree}
is.binary.tree(tree) #FALSE (i.e. there are polytomies)
is.ultrametric(tree) #the tree has no branch lengths
```

Polytomies are present within and between Acer Betula Quercus Fraxinus.

```{r resolve polytomies at random}
set.seed(2017) #making it replicable
tree_random <- multi2di(tree,random=TRUE)
is.binary.tree(tree_random) #TRUE
```

Plot the tree:
```{r plot tree_random, fig.width=10, fig.height=5, echo=TRUE, message=FALSE}
plot(tree_random)
```


#Make sure that species names match exactly tree tip.labals